---
title: "Lecture 15"
output: html_document
---

## Lecture 15

Categorical Predictors and Interactions

T

```{r}
1 + 1

```

An **outlier** is a point that has a response value ($y_i$) that does not follow the trend set by the data

We can classify a point as an outlier if its standardized residual falls outside the interval from **-2 to 2**. For large data sets, change this rule **-4 to 4** (otherwise, too many points would be flagged).

Just because a point is an outlier and / or has high leverage does not mean we must ignore that point and remove it from the model. Rather, outliers and / or high leverage points should be investigated, and can provide important insights about the data. Sometimes outliers and / or leverage points indicate a problem with the data that can be corrected.

Residual plots are one of the most useful diagnostics for a multiple linear regression model.

The most important diagnostic is a plot of the residuals, $\hat e_i$, versus the fitted values, $\hat y_i$. Alternatively, we can use the standardized residuals, $r_i$, instead of the raw residuals.

Ideally, the residual plots should show no obvious patterns or nonconstant variability, and the points are randomly scattered around 0.

Recall, the data set from Zagat surveys of customers of 168 Italian restaurants in New York City. We considered the following multiple linear regression model:

$$
Y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_3+\epsilon
$$

```{r}
nyc <- read.csv("nyc.csv")

head(nyc)
```

```{r}
library(ISLR)

head(Credit)
```

The response variable is Balance, the average credit card balance in dollars

The predictors of interest are Income and Student a dummy variable indicating student status

```{r}
lm1 <- lm(Balance ~ Income + Student, data=Credit)

contrasts(Credit$Student)
```

```{r}
summary(lm1)
```
